Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/evaluation.tex
add para about recursion, and figure demonstrating jump compression

\paragraph{Jump Compression}
%
A \emph{jump compressed} trace is one whose edges are limited to forward
or backward jumps. (The sizes of these traces is equal to the jump
metric below.)
%
In our experience, jump compression abstracts many details of the
computation that are often uninteresting or irrelevant to the
explanation, or at any rate, easily summarized as call-return
pairs.
%
Expanding the initial visualization to a jump-compressed trace is often
an ideal first step;
% expanding a jump-compressed traces ideal for the initial state of the
they provide a concise overview of the execution and allow the user to
step directly into an interesting function call, at which point the can
proceed with more fine-grained steps.
%
Note that jump compressed traces are not quite the same as
stack-traces as they show \emph{all} function calls, including
those that returned successfully.
%
For example, the display on the right in Figure~\ref{fig:factorial} shows
a jump-compressed trace, of size 4. The full trace for
the same example has size 19.

>>>>

\paragraph{Jump Compression}
%
A \emph{jump compressed} trace is one whose edges are limited to forward
or backward jumps. (The sizes of these traces is equal to the jump
metric below.)
%
In our experience, jump compression abstracts many details of the
computation that are often uninteresting or irrelevant to the
explanation.
% 
In particular, jump compressed traces hide low-level operations and
summarize function calls as call-return pairs, see
Figure~\ref{fig:fac-jump} for a variant of @fac@ that implements the
subtraction as a function call instead of a primitive.
%
\begin{figure}[t]
\centering
\includegraphics[height=1.5in]{fac-minus.png}
\caption{Jump-compressed trace of \texttt{fac 1} with subtraction
  implemented as a function call.}
\label{fig:fac-jump}
\end{figure}
%
Once the user has identified an interesting call-return pair, they can
step into the call and proceed with more fine-grained steps.
%
% Expanding the initial visualization to a jump-compressed trace is often
% an ideal first step;
% % expanding a jump-compressed traces ideal for the initial state of the
% they provide a concise overview of the execution and allow the user to
% step directly into an interesting function call, at which point the can
% proceed with more fine-grained steps.
%
Note that jump compressed traces are not quite the same as
stack-traces as they show \emph{all} function calls, including
those that returned successfully.
%
For example, the display on the right in Figure~\ref{fig:factorial} shows
a jump-compressed trace, of size 4. The full trace for
the same example has size 19.

### 2015-11-20 paper/pldi16-submission/related.tex
add para about recursion, and figure demonstrating jump compression

\subsection{Type Errors}
\label{sec:type-error}

\paragraph{Diagnosis and Repair}
\label{sec:diagnosis-repair}
It is well known that unification-based type inference procedures can
produce poor error messages, and in particular, can misidentify the
\emph{source} of the type error.
%
Thus, many groups have explored techniques to pinpoint the true source
of the error and recommend fixes.

@@

\subsection{Testing}
\label{sec:testing}
\nanomaly is at its heart a test generator, and builds on a rich line of
work.
%
Our use of holes to represent unknown values is inspired by the work of
Runciman, Naylor, and Lindblad~\cite{runciman_smallcheck_2008,naylor_finding_2007,lindblad_property_2007},
%
who use lazy evaluation to drastically reduce the search space for
exhaustive test generation, by grouping together equivalent inputs by
the set of values they force. An exhaustive search is complete (up to
the depth bound), if a witness exists it will be found, but due to the
exponential blowup in the search space the depth bound can be quite
limited without advanced grouping and filtering techniques.
%
Our search is not exhaustive; instead we use random generation to fill
in holes on demand.
%
Random test generation~\cite{claessen_quickcheck:_2000,csallner_jcrasher:_2004,pacheco_feedback-directed_2007}
%
is by its nature incomplete, but is able to check larger inputs than
exhaustive testing as a result.

@@

\subsection{Program Exploration}
Perera \etal~\cite{perera_functional_2012} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} -- containing holes -- and present a partial slice,
containing only those steps that affected the the partial value. This
system is designed to answer questions of the form ``Where did this
value come from?'' and thus is focused on backward exploration. In
contrast, our trace visualization supports forward \emph{and} backward
exploration, though our backward steps are more limited than
Perera's. Specifically, we do not support selecting a value and
inserting the intermediate terms that preceded it while ignoring
unrelated computation steps; this would be interesting future work.

>>>>

% \paragraph{Type Errors}
% \label{sec:type-error}

\paragraph{Diagnosing and Repairing Type Errors}
\label{sec:diagnosis-repair}
It is well known that unification-based type inference procedures can
produce poor error messages, and in particular, can misidentify the
\emph{source} of the type error.
%
Thus, many groups have explored techniques to pinpoint the true source
of the error and recommend fixes.

@@

\paragraph{Testing}
\label{sec:testing}
\nanomaly is at its heart a test generator, and builds on a rich line of
work.
%
Our use of holes to represent unknown values is inspired by the work of
Runciman, Naylor, and Lindblad~\cite{runciman_smallcheck_2008,naylor_finding_2007,lindblad_property_2007},
%
who use lazy evaluation to drastically reduce the search space for
exhaustive test generation, by grouping together equivalent inputs by
the set of values they force. An exhaustive search is complete (up to
the depth bound), if a witness exists it will be found, but due to the
exponential blowup in the search space the depth bound can be quite
limited without advanced grouping and filtering techniques.
%
Our search is not exhaustive; instead we use random generation to fill
in holes on demand.
%
Random test generation~\cite{claessen_quickcheck:_2000,csallner_jcrasher:_2004,pacheco_feedback-directed_2007}
%
is by its nature incomplete, but is able to check larger inputs than
exhaustive testing as a result.

@@

\paragraph{Program Exploration}
Perera \etal~\cite{perera_functional_2012} present a tracing semantics
for functional programs that tags values with their provenance, enabling
a form of backwards program slicing from a final value to the sequence
of reductions that produced it. Notably, they allow the user to supply a
\emph{partial value} -- containing holes -- and present a partial slice,
containing only those steps that affected the the partial value. This
system is designed to answer questions of the form ``Where did this
value come from?'' and thus is focused on backward exploration. In
contrast, our trace visualization supports forward \emph{and} backward
exploration, though our backward steps are more limited than
Perera's. Specifically, we do not support selecting a value and
inserting the intermediate terms that preceded it while ignoring
unrelated computation steps; this would be interesting future work.

### 2015-11-20 paper/pldi16-submission/semantics.tex
add para about recursion, and figure demonstrating jump compression

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and an empty value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda, it may not map \emph{directly} to another hole, so in the second
case above we do not need to \forcesym\ the result of the substitution.

@@

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, and
function application --- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

>>>>

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and an empty value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda, it may not map \emph{directly} to another hole, so in the second
case above we do not need to \forcesym\ the result of the substitution,
the sub-hole will be narrowed when the context demands it.

@@

\paragraph{Primitive Reductions}
Primitive reduction steps --- addition, if-elimination, and
function application --- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that $v_1$ is a function.

\paragraph{Recursion}
Fixed-point operators often cannot be typed in static type systems, but
we are not concerned with \emph{assigning} types to terms, rather with
showing that \emph{no type} can be assigned. We are simply executing the
untyped $\lambda$-calculus, which has no issue handling recursion.
